---
output: 
  pdf_document:
    latex_engine: xelatex
    number_sections: true
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Appendix 2 – Analytical Methods and Code {-}

## Overview

This document contains the reproducible R code used for data discovery, cleaning, validation, exploratory analysis, and statistical modelling of breast cancer screening uptake across 12 clinical team datasets. Scripts are presented in the order executed.

## Computational Environment

The following R packages were used in the analysis (versions shown below):
```{r package-versions, echo=FALSE}
packages_used <- c(
  "tidyverse",
  "naniar",
  "ComplexUpset",
  "effectsize",
  "binom",
  "broom",
  "scales",
  "janitor",
  "skimr",
  "car",
  "sessioninfo"
)

sapply(packages_used, function(p) {
  as.character(packageVersion(p))
})
```

All analyses were conducted in R (version 4.5.2, 2025-10-31 ucrt) using RStudio (version 2026.01.0+392) on a Windows 10 x64 (build 19045) operating system.

## Project Directory Structure

```{r project-structure, echo=FALSE}
cat("
Costello_Medical_Screening_Analysis/
├── raw_data/
│   ├── DN01.csv
│   ├── DN02.csv
│   └── ...
├── output/
│   ├── breast_screening_cleaned.csv
│   ├── Missing_values_per_variable.png
│   ├── number_of_dependents_by_clinical_team.png
│   └── ...
├── scripts/
│   ├── 01_data_discovery.R
│   ├── 02_data_cleaning.R
│   ├── 03_validation.R
│   ├── 04_exploratory_data_analysis.R
│   ├── 05_statistical_modelling.R
│   └── Costello Medical Screening Analysis in R.Rmd
├── report/
│   ├── Report.docx
│   └── Costello-Medical-Screening-Analysis-in-R.pdf
└── Costello_Medical_Screening_Analysis.Rproj
")
```

## Data Discovery (01_data_discovery.R)

```{r data-discovery, echo=TRUE, eval=FALSE}
# ==========================================================
# Script: 01_data_discovery.R
# Purpose: Initial exploration of raw screening datasets 
#          prior to cleaning and harmonisation.
# Author: Nikolaos Samperis
# Date: 20/02/2026
# ==========================================================

# ----------------------------------------------------------
# 1. Load required libraries
# ----------------------------------------------------------
# tidyverse: data manipulation and iteration
# stringr:   numeric-aware sorting of file names

library(tidyverse)
library(stringr)

# ----------------------------------------------------------
# 2. Identify and order raw CSV files
# ----------------------------------------------------------
# Assumes all raw datasets are stored in a folder named 
# "raw_data". Files are sorted numerically to preserve the 
# intended dataset order (e.g., DN01, DN02, ...).

files <- list.files("raw_data",
                    pattern = "\\.csv$",
                    full.names = TRUE) |>
  str_sort(numeric = TRUE)

# ----------------------------------------------------------
# 3. Import all datasets into a named list
# ----------------------------------------------------------
# Each dataset is read as a tibble.
# Names are assigned using the file names for traceability.

datasets <- set_names(files, basename(files)) |>
  map(~ read_csv(.x, show_col_types = FALSE))

# ----------------------------------------------------------
# 4. Structural inspection of each dataset
# ----------------------------------------------------------
# Displays variable names, types, and preview of values.
# Used to detect inconsistencies in structure across files.

walk2(datasets, names(datasets), ~ {
  cat("\n\n=========================\n")
  cat("Dataset:", .y, "\n")
  cat("=========================\n")
  glimpse(.x)
})

# ----------------------------------------------------------
# 5. Check variable names across datasets
# ----------------------------------------------------------
# Used to identify naming inconsistencies prior to cleaning.

map(datasets, names)

# ----------------------------------------------------------
# 6. Check variable data types
# ----------------------------------------------------------
# Identifies potential inconsistencies (e.g., numeric vs character)
# across datasets that may require harmonisation.

map_dfr(datasets,
        ~ summarise_all(.x, class),
        .id = "dataset")

# ----------------------------------------------------------
# 7. Dataset-level dimensions and missingness
# ----------------------------------------------------------
# For each dataset:
#   - number of rows
#   - number of columns
#   - total number of missing values
#   - percentage of missing values
# This provides an overview of completeness by file.

map_dfr(datasets, ~ tibble(
  rows = nrow(.x),
  cols = ncol(.x),
  total_missing = sum(is.na(.x)),
  missing_pct = mean(is.na(as.matrix(.x))) * 100
), .id = "dataset")

# ----------------------------------------------------------
# 8. Missing values per variable within each dataset
# ----------------------------------------------------------
# Summarises number of missing values per column for each dataset.
# Used to detect systematic missingness patterns.

map_dfr(datasets, ~
          summarise(.x,
                    across(everything(),
                           ~ sum(is.na(.)))),
        .id = "dataset")

# ----------------------------------------------------------
# 9. Identify duplicated rows within each dataset
# ----------------------------------------------------------
# Flags fully duplicated records (forward or backward).
# No removal is performed at this stage — this is exploratory.

walk2(datasets, names(datasets), ~ {
  
  dup_rows <- .x[duplicated(.x) | duplicated(.x, fromLast = TRUE), ]
  
  if(nrow(dup_rows) > 0){
    cat("\nDuplicated rows in:", .y, "\n")
    print(dup_rows)
  }
})

# ----------------------------------------------------------
# 10. Summary statistics for numeric variables
# ----------------------------------------------------------
# Calculates min, median, mean, and max for numeric variables
# in each dataset.
# Used to detect:
#   - Implausible values
#   - Data entry errors
#   - Outliers
#   - Distribution differences between datasets

map_dfr(datasets, ~
          summarise(.x,
                    across(where(is.numeric),
                           list(
                             min = ~ min(.x, na.rm = TRUE),
                             median = ~ median(.x, na.rm = TRUE),
                             mean = ~ mean(.x, na.rm = TRUE),
                             max = ~ max(.x, na.rm = TRUE)
                           ))),
        .id = "dataset")





```

## Data Cleaning and Harmonisation (02_data_cleaning.R)

```{r data-cleaning, echo=TRUE, eval=FALSE}
# ==========================================================
# Script: 02_data_cleaning.R
# Purpose: Clean, harmonise, and combine raw screening datasets
#          into a single analysis-ready dataset.
# Author: Nikolaos Samperis
# Date: 20/02/2026
# ==========================================================


# ----------------------------------------------------------
# 1. Load required libraries
# ----------------------------------------------------------
# tidyverse: data manipulation
# stringr:   file name processing

library(tidyverse)
library(stringr)

# ----------------------------------------------------------
# 2. Load and order raw datasets
# ----------------------------------------------------------
# Ensures consistent dataset ordering (DN01–DN012).

files <- list.files("raw_data",
                    pattern = "\\.csv$",
                    full.names = TRUE) |>
  str_sort(numeric = TRUE)

datasets <- set_names(files, basename(files)) |>
  map(~ read_csv(.x, show_col_types = FALSE))

# ----------------------------------------------------------
# 3. Remove auto-generated index column
# ----------------------------------------------------------
# Some CSV files contain an export index column ("...1").
# This is not analytically meaningful and is removed.

datasets <- map(datasets, ~ {
  if(names(.x)[1] == "...1") {
    select(.x, -1)
  } else {
    .x
  }
})

# ----------------------------------------------------------
# 4. Standardise variable names
# ----------------------------------------------------------
# Variable names are harmonised across datasets to ensure
# consistent naming prior to merging.

datasets <- map(datasets, ~ {
  
  df <- .x
  names_lower <- tolower(names(df))
  
  # Rename based on lowercase comparison
  if("screening" %in% names_lower)
    df <- rename(df, screening_decision = !!names(df)[which(names_lower == "screening")])
  
  if("age" %in% names_lower)
    df <- rename(df, age = !!names(df)[which(names_lower == "age")])
  
  if("weight" %in% names_lower)
    df <- rename(df, weight = !!names(df)[which(names_lower == "weight")])
  
  if("height" %in% names_lower)
    df <- rename(df, height = !!names(df)[which(names_lower == "height")])
  
  if("alcohol" %in% names_lower)
    df <- rename(df, alcohol_units = !!names(df)[which(names_lower == "alcohol")])
  
  if("es" %in% names_lower)
    df <- rename(df, employment_status = !!names(df)[which(names_lower == "es")])
  
  if("deps" %in% names_lower)
    df <- rename(df, n_dependents = !!names(df)[which(names_lower == "deps")])
  
  df
})

# ----------------------------------------------------------
# 5. Ensure structural consistency (DN01 issue)
# ----------------------------------------------------------
# DN01 does not contain employment_status.
# Add missing column to preserve consistent structure.

datasets <- map(datasets, ~ {
  if(!"employment_status" %in% names(.x)) {
    .x$employment_status <- NA
  }
  .x
})

# ----------------------------------------------------------
# 6. Standardise column order
# ----------------------------------------------------------
# Enforces identical column ordering across datasets.

standard_cols <- c(
  "age",
  "weight",
  "height",
  "alcohol_units",
  "employment_status",
  "n_dependents",
  "screening_decision"
)

datasets <- map(datasets, ~ .x[, standard_cols])

# ----------------------------------------------------------
# 7. Type conversion and unit harmonisation
# ----------------------------------------------------------
# Convert variables to appropriate types.
# Height converted from centimetres to metres for analysis.

datasets <- map(datasets, ~ {
  
  .x |>
    mutate(
      age = as.numeric(age),
      weight = as.numeric(weight),
      height = as.numeric(height) / 100,   # cm → meters (no rounding)
      alcohol_units = as.numeric(alcohol_units),
      n_dependents = as.integer(n_dependents),
      employment_status = as.factor(employment_status),
      screening_decision = as.factor(screening_decision)
    )
})

# ----------------------------------------------------------
# 8. Add clinical team identifier and merge
# ----------------------------------------------------------
# Adds dataset source identifier before row-binding.


combined_data <- imap_dfr(
  datasets,
  ~ mutate(.x,
           clinical_team = str_remove(.y, "\\.csv$"))
)


combined_data$clinical_team <- factor(combined_data$clinical_team)

# ----------------------------------------------------------
# 9. Final validation checks
# ----------------------------------------------------------
# Verify structure and basic summaries of merged dataset.

glimpse(combined_data)
summary(combined_data)

# Confirm height now expressed in metres
summary(combined_data$height)

# ----------------------------------------------------------
# 10. Export cleaned dataset
# ----------------------------------------------------------
# Round numeric display precision at export stage only
# (avoids floating-point artefacts while preserving accuracy).

combined_data <- combined_data |>
  mutate(
    height = round(height, 3),
    weight = round(weight, 3)
  )

if(!dir.exists("output")) {
  dir.create("output")
}

write_csv(combined_data,
          file = "output/breast_screening_cleaned.csv")
```

## Data Validation (03_validation.R)

```{r validation_checks, echo=TRUE, eval=FALSE}
# ==========================================================
# Script: 03_validation.R
# Purpose: Perform structural and logical validation checks
#          on the cleaned breast screening dataset.
# Author: Nikolaos Samperis
# Date: 20/02/2026
# ==========================================================

# ----------------------------------------------------------
# 1. Load required libraries and dataset
# ----------------------------------------------------------
# tidyverse: data manipulation
# janitor:   column name validation
# skimr:     structured dataset summary

library(tidyverse)
library(janitor)
library(skimr)

# Import cleaned dataset with explicit column types
# (ensures structural consistency at validation stage)

data <- read_csv(
  "output/breast_screening_cleaned.csv",
  col_types = cols(
    age = col_double(),
    weight = col_double(),
    height = col_double(),
    alcohol_units = col_double(),
    n_dependents = col_integer(),
    employment_status = col_factor(),
    screening_decision = col_factor(),
    clinical_team = col_factor()
  )
)

cat("Validation started...\n\n")

# ----------------------------------------------------------
# 2. Structural validation
# ----------------------------------------------------------
# Confirm dataset dimensions and column integrity.

cat("===== STRUCTURE =====\n")
cat("Rows:", nrow(data), "\n")
cat("Columns:", ncol(data), "\n\n")

# Ensure column names follow clean naming conventions
if(any(names(data) != make_clean_names(names(data)))) {
  warning("Column names are not fully standardised.")
}

# ----------------------------------------------------------
# 3. Duplicate record assessment
# ----------------------------------------------------------
# Identify fully duplicated rows (including first occurrence).

cat("===== DUPLICATE CHECK =====\n")

duplicate_rows <- data[
  duplicated(data) | duplicated(data, fromLast = TRUE),
]

duplicate_count <- nrow(duplicate_rows)

cat("Number of duplicated rows:", duplicate_count, "\n")

if(duplicate_count == 0){
  cat("No duplicated rows found.\n\n")
} else {
  cat("Duplicated rows detected. Displaying all duplicates:\n")
  print(duplicate_rows)
  cat("\n")
}

# ----------------------------------------------------------
# 4. Missing data validation
# ----------------------------------------------------------
# Assess missingness overall and by clinical team
# to detect potential systematic patterns.

if("clinical_team" %in% names(data)){
  
  cat("===== MISSINGNESS BY CLINICAL TEAM =====\n")
  
  vars_to_check <- setdiff(names(data), "clinical_team")
  
  missing_by_team <- data %>%
    group_by(clinical_team) %>%
    summarise(
      across(
        all_of(vars_to_check),
        list(
          missing_count  = ~ sum(is.na(.)),
          missing_percent = ~ round(mean(is.na(.)) * 100, 2)
        )
      ),
      .groups = "drop"
    ) %>%
    pivot_longer(
      cols = -clinical_team,
      names_to = c("variable", ".value"),
      names_pattern = "(.+)_(missing_count|missing_percent)"
    ) %>%
    filter(missing_count > 0) %>%
    arrange(desc(missing_percent))
  
  if(nrow(missing_by_team) == 0){
    cat("No missing values detected across clinical teams.\n\n")
  } else {
    print(missing_by_team, n = Inf)
    cat("\n")
  }
}

# ----------------------------------------------------------
# 5. Logical plausibility checks
# ----------------------------------------------------------
# Evaluate whether key numeric variables fall within
# clinically plausible ranges.

cat("===== LOGICAL CHECKS =====\n")

# Age plausibility (adult screening population)
if("age" %in% names(data)){
  implausible_age <- sum(data$age < 18 | data$age > 100, na.rm = TRUE)
  cat("Implausible age values (<18 or >100):", implausible_age, "\n")
}

# Height plausibility (in metres)
if("height" %in% names(data)){
  implausible_height <- sum(data$height < 1.3 | data$height > 2.2, na.rm = TRUE)
  cat("Implausible height values (<1.3m or >2.2m):", implausible_height, "\n")
}

# Weight plausibility (in kg)
if("weight" %in% names(data)){
  implausible_weight <- sum(data$weight < 30 | data$weight > 250, na.rm = TRUE)
  cat("Implausible weight values (<30kg or >250kg):", implausible_weight, "\n")
}

cat("\n")

# ----------------------------------------------------------
# 6. Overall dataset profile
# ----------------------------------------------------------
# Generate structured summary of variable distributions
# to support exploratory interpretation.

cat("===== SKIM SUMMARY =====\n")
skim(data)

cat("\nValidation completed successfully.\n")
```

## Exploratory Data Analysis (04_exploratory_data_analysis.R)

```{r EDA, echo=TRUE, eval=FALSE}
# ==========================================================
# Script: 04_exploratory_analysis.R
# Purpose: Perform exploratory and descriptive analysis of
#          the cleaned breast screening dataset.
# Author: Nikolaos Samperis
# Date: 20/02/2026
# ==========================================================


# ----------------------------------------------------------
# 1. Load required libraries
# ----------------------------------------------------------
# tidyverse: data manipulation & visualisation
# naniar: missing data visualisation
# ComplexUpset: missingness pattern plots
# effectsize: effect size calculations
# binom: binomial confidence intervals
# broom: model tidying

library(tidyverse)
library(naniar)
library(ComplexUpset)
library(effectsize)
library(binom)
library(broom)

# ----------------------------------------------------------
# 2. Load cleaned dataset
# ----------------------------------------------------------

# Ensure categorical variables are correctly specified
data <- data |>
  mutate(
    clinical_team = factor(clinical_team),
    screening_decision = factor(screening_decision),
    employment_status = factor(employment_status)
  )

# ----------------------------------------------------------
# 3. Overall missingness assessment
# ----------------------------------------------------------
# Summarise total missing values per variable.

missing_summary <- data.frame(
  variable = names(data),
  missing_count = colSums(is.na(data)),
  missing_percent = round(colMeans(is.na(data)) * 100, 2)
) |>
  arrange(desc(missing_percent))

print("===== OVERALL MISSINGNESS =====")
print(missing_summary)

# Visualisation of missingness per variable
p <- gg_miss_var(data)

for (i in seq_along(p$layers)) {
  g <- class(p$layers[[i]]$geom)[1]
  
  if (g == "GeomPoint") {
    p$layers[[i]]$aes_params$size <- 3
  }
  
  if (g == "GeomBar") {
    p$layers[[i]]$aes_params$linewidth <- 1.5
    p$layers[[i]]$aes_params$size <- 1.5
  }
}

print(
  p +
    labs(
      title = "Missing Values per Recorded Patient Characteristic",
      y = "Number of Missing Observations",
      x = NULL
    ) +
    theme(
      plot.title = element_text(face = "bold", margin = margin(b = 20), hjust = 0.5),
      axis.title.x = element_text(margin = margin(t = 20)),
      legend.position = "none"
    )
)

# ----------------------------------------------------------
# 4. Missingness pattern (UpSet)
# ----------------------------------------------------------
# Examine joint missingness patterns across key variables.

data_upset <- data |>
  mutate(
    missing_alcohol_units     = is.na(alcohol_units),
    missing_employment_status = is.na(employment_status),
    missing_height            = is.na(height)
  )

upset(
  data_upset,
  intersect = c(
    "missing_alcohol_units",
    "missing_employment_status",
    "missing_height"
  ),
  name = "Missing Pattern",
  
  base_annotations = list(
    'Intersection size' =
      intersection_size(text = list(size = 3)) +
      labs(y = "Intersection size") +
      ggtitle("Patterns of Missing Data Across Patient Characteristics") +
      theme(
        plot.title = element_text(
          face = "bold",
          hjust = 0.5,
          margin = margin(b = 15)
        ),
        plot.margin = margin(t = 5, r = 5, b = 5, l = 0)
      )
  ),
  
  set_sizes =
    upset_set_size() +
    labs(caption = "Total missing per variable") +
    theme(
      plot.caption = element_text(
        hjust = 0.5,
        vjust = 7,
        size = 11
      ),
      axis.title.x = element_blank(),
      axis.title.y = element_blank()
    )
) +
  labs(x = "Missingness Combination") +
  theme(
    axis.title.x = element_text(margin = margin(t = 15))
  )

# ----------------------------------------------------------
# 5. Missingness by clinical team
# ----------------------------------------------------------
# Calculate percentage missing per variable within each team.

missing_by_team <- data %>%
  group_by(clinical_team) %>%
  summarise(
    alcohol_missing = mean(is.na(alcohol_units)) * 100,
    height_missing  = mean(is.na(height)) * 100,
    weight_missing  = mean(is.na(weight)) * 100,
    employment_missing = mean(is.na(employment_status)) * 100
  )

print("===== MISSINGNESS BY CLINICAL TEAM (%) =====")
print(missing_by_team)

# Set clinical_team levels to numerical order (DN01–DN12)
# Ensures consistent ordering in plots and summaries.

data$clinical_team <- factor(
  data$clinical_team,
  levels = c("DN01","DN02","DN03","DN04","DN05","DN06",
             "DN07","DN08","DN09","DN010","DN011","DN012")
)


# Plot alcohol missingness proportion by clinical team
ggplot(data, aes(x = clinical_team, fill = is.na(alcohol_units))) +
  geom_bar(position = "fill") +
  labs(
    title = "Alcohol Intake Missingness by Clinical Team",
    x = "Clinical Team",
    y = "Proportion",
    fill = "Alcohol Missing"
  ) +
  theme(
    plot.title = element_text(
      face = "bold",
      hjust = 0.5,               # center title
      margin = margin(b = 15)    # space below title
    ),
    axis.title.x = element_text(
      margin = margin(t = 15)    # space between x label and plot
    ),
    axis.title.y = element_text(
      margin = margin(r = 15)    # space between y label and plot
    )
  )

# ----------------------------------------------------------
# 6. Assessment of missingness mechanism
# ----------------------------------------------------------
# Evaluate associations between missingness indicators and
# screening behaviour or demographic variables.

# --------- Alcohol missing indicator ----------
data$alcohol_missing <- is.na(data$alcohol_units)

# Association with screening decision
print("===== CHI-SQUARE: Alcohol Missing vs Screening =====")
print(table(data$screening_decision, data$alcohol_missing))
print(chisq.test(data$screening_decision, data$alcohol_missing))

# Association with age
print("===== T-TEST: Age vs Alcohol Missing =====")
print(t.test(age ~ alcohol_missing, data = data))

# ------ Height missing indicator ------
data$height_missing <- is.na(data$height)

print("===== FISHER'S EXACT TEST: Height Missing vs Screening decision =====")
tab_height <- table(data$screening_decision, data$height_missing)
print(tab_height)
print(fisher.test(tab_height))

print("===== T-TEST: Age vs Height Missing =====")
print(t.test(age ~ height_missing, data = data))

# ---------- Employment missing indicator -------------
data$employment_missing <- is.na(data$employment_status)

print("===== CHI-SQUARE: Employment Missing vs Screening decision =====")
print(table(data$screening_decision, data$employment_missing))
print(chisq.test(data$screening_decision, data$employment_missing))

# ----------------------------------------------------------
# 7. Visual comparison of age distribution by missingness
# ----------------------------------------------------------
# Compare age density distributions between individuals
# with and without missing alcohol intake data.

ggplot(data, aes(x = age, fill = alcohol_missing)) +
  geom_density(alpha = 0.4) +
  labs(
    title = "Age Distribution by Alcohol Missingness",
    x = "Age (years)",
    y = "Density",
    fill = "Alcohol Missing"
  ) +
  theme(
    plot.title = element_text(
      face = "bold",
      hjust = 0.5,              # center title
      margin = margin(b = 15)   # space below title
    ),
    axis.title.x = element_text(
      margin = margin(t = 15)   # space between x label and panel
    ),
    axis.title.y = element_text(
      margin = margin(r = 15)   # space between y label and panel
    )
  )


# ----------------------------------------------------------
# 8. Association Between Missingness and Clinical Team
# ----------------------------------------------------------
# Test whether missingness varies systematically across
# clinical teams using chi-square tests.

test_missing_vs_team <- function(var_name) {
  
  missing_indicator <- is.na(data[[var_name]])
  
  cat("\n====================================\n")
  cat("Missingness vs Clinical Team:", var_name, "\n")
  cat("====================================\n")
  
  tab <- table(data$clinical_team, missing_indicator)
  print(tab)
  
  # Chi-square test for 12x2 contingency table
  print(chisq.test(tab))
}


# Run tests for selected variables
test_missing_vs_team("alcohol_units")
test_missing_vs_team("height")
test_missing_vs_team("employment_status")

# ----------------------------------------------------------
# 9. Summary Statistics by Clinical Team (Continuous Variables)
# ----------------------------------------------------------
# Calculate mean, SD, median, and IQR for numeric variables
# within each clinical team.

team_summary <- data %>%
  group_by(clinical_team) %>%
  summarise(
    n = n(),
    across(
      c(age, weight, height, alcohol_units, n_dependents),
      list(
        mean = ~mean(.x, na.rm = TRUE),
        sd = ~sd(.x, na.rm = TRUE),
        median = ~median(.x, na.rm = TRUE),
        IQR = ~IQR(.x, na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  )

team_summary

# ----------------------------------------------------------
# 10. Distribution of Continuous Variables by Clinical Team
# ----------------------------------------------------------
# Visual inspection of within-team distributions using
# faceted histograms.

plot_histogram_by_team_facet <- function(df, var_name, bins = 20) {
  
  ggplot(df, aes(x = .data[[var_name]])) +
    geom_histogram(
      bins = bins,
      fill = "steelblue",
      colour = "black",
      na.rm = TRUE
    ) +
    facet_wrap(~ clinical_team) +
    labs(
      title = paste("Distribution of", var_name, "by Clinical Team"),
      x = var_name,
      y = "Frequency"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5)
    )
}

numeric_vars <- c("age", "weight", "height", "alcohol_units")

for (var in numeric_vars) {
  print(plot_histogram_by_team_facet(data, var))
}

# ------------------------------------------------------------
# 11. Descriptive Analysis of Categorical & Discrete Variables
# ------------------------------------------------------------

# Employment status distribution within each team
employment_by_team <- data %>%
  count(clinical_team, employment_status) %>%
  group_by(clinical_team) %>%
  mutate(
    percent = round(100 * n / sum(n), 1)
  ) %>%
  ungroup()

employment_by_team

ggplot(data, aes(x = employment_status)) +
  geom_bar(fill = "steelblue", colour = "black") +
  facet_wrap(~ clinical_team) +
  labs(
    title = "Employment Status Distribution by Clinical Team",
    x = "Employment Status",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


# Screening decision distribution within each team
screening_by_team <- data %>%
  count(clinical_team, screening_decision) %>%
  group_by(clinical_team) %>%
  mutate(
    percent = round(100 * n / sum(n), 1)
  ) %>%
  ungroup()

screening_by_team

ggplot(data, aes(x = screening_decision)) +
  geom_bar(fill = "steelblue", colour = "black") +
  facet_wrap(~ clinical_team) +
  labs(
    title = "Screening Decision by Clinical Team",
    x = "Screening Decision",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  )


# Number of dependents distribution within each team
dependents_by_team <- data %>%
  count(clinical_team, n_dependents) %>%
  group_by(clinical_team) %>%
  mutate(
    percent = round(100 * n / sum(n), 1)
  ) %>%
  ungroup()

dependents_by_team

# Bar chart: Distribution of number of dependents by clinical team
# Treated as categorical for clearer visual comparison
ggplot(data, aes(x = factor(n_dependents))) +
  geom_bar(fill = "steelblue", colour = "black") +
  facet_wrap(~ clinical_team) +
  labs(
    title = "Number of Dependents by Clinical Team",
    x = "Number of Dependents",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

# ----------------------------------------------------------
# 12. Between-Team Heterogeneity (Continuous Variables)
# ----------------------------------------------------------
# Assess differences across clinical teams using ANOVA
# and non-parametric testing where appropriate.

continuous_vars <- c("age", "weight", "height", "alcohol_units")

anova_results <- lapply(continuous_vars, function(var) {
  formula <- as.formula(paste(var, "~ clinical_team"))
  summary(aov(formula, data = data))
})

names(anova_results) <- continuous_vars
anova_results

# Kruskal–Wallis test for alcohol intake across clinical teams
# as it appears moderately right-skewed
kruskal_alcohol <- kruskal.test(alcohol_units ~ clinical_team, data = data)

kruskal_alcohol

# Height showed significant between-team heterogeneity (ANOVA p < 0.001).
# DN012 had a notably higher mean height compared to other teams (~1.75m vs ~1.60m),
# raising the possibility of a unit or scaling inconsistency.
# Raw DN012 values are inspected to verify measurement units 
# and rule out harmonisation errors.

# Raw DN012 heights
summary(read.csv("raw_data/DN012.csv")$Height)

# Cleaned DN012 heights
data %>%
  filter(clinical_team == "DN012") %>%
  summarise(min = min(height, na.rm = TRUE),
            max = max(height, na.rm = TRUE),
            mean = mean(height, na.rm = TRUE))

# Raw DN012 heights were recorded in centimetres and correctly converted 
# to metres during cleaning. No scaling inconsistencies detected.
# Observed between-team heterogeneity in height reflects genuine 
# demographic variation rather than a data processing error.

# Effect size (eta-squared) for height differences
model_height <- aov(height ~ clinical_team, data = data)
eta_height <- eta_squared(model_height)
eta_height

# ----------------------------------------------------------
# 13. Between-Team Heterogeneity (Categorical Variables)
# ----------------------------------------------------------
# Use chi-square tests to assess variation across teams.

# Exclude DN01 (employment not recorded) and drop unused factor levels
# to ensure valid chi-square testing.
data_noDN01 <- droplevels(
  data[data$clinical_team != "DN01", ]
)

# Contingency table
table_employment <- table(
  data_noDN01$employment_status,
  data_noDN01$clinical_team
)
table_employment

# Chi-square test
chisq.test(table_employment)

# Contingency table
table_screening <- table(data$screening_decision, data$clinical_team)
table_screening

# Chi-square test
chisq.test(table_screening)


# Contingency table
table_dependents <- table(data$n_dependents, data$clinical_team)
table_dependents

# Chi-square test
chisq.test(table_dependents)


# Although screening uptake showed statistical heterogeneity across teams,
# effect size is required to assess the practical magnitude of this association.
# Cramér’s V provides a standardized measure of association for categorical data.
cramers_v(table_screening)

# Calculate within-team screening proportions to examine differences
# in uptake rates across clinical teams.
prop.table(table_screening, margin = 2)

# ----------------------------------------------------------
# 14. Screening Uptake by Clinical Team
# ----------------------------------------------------------
# Estimate screening proportions with Wilson 95% confidence intervals.

screening_plot_data <- data %>%
  group_by(clinical_team) %>%
  summarise(
    total_n = n(),
    n_positive = sum(screening_decision == "Y"),
    prop_screened = n_positive / total_n,
    .groups = "drop"
  )

ci <- binom.confint(
  screening_plot_data$n_positive,
  screening_plot_data$total_n,
  methods = "wilson"
)

screening_plot_data$lower_ci <- ci$lower
screening_plot_data$upper_ci <- ci$upper

# Calculate overall mean screening acceptance across teams
overall_mean <- mean(screening_plot_data$prop_screened)

# Preserve ascending order and include sample size in x-axis label
screening_plot_data <- screening_plot_data %>%
  arrange(clinical_team) %>%
  mutate(
    team_label = paste0(clinical_team, "\n(n=", total_n, ")"),
    team_label = factor(team_label, levels = team_label)
  )

# Create bar plot of screening uptake by clinical team
ggplot(screening_plot_data, aes(x = team_label, y = prop_screened)) +
  
  # Bars representing screening acceptance per team
  geom_col(fill = "steelblue") +
  
  # Add 95% binomial confidence intervals
  geom_errorbar(
    aes(ymin = lower_ci, ymax = upper_ci),
    width = 0.2,
    linewidth = 0.4
  ) +
  
  # Add percentage labels above bars
  geom_text(
    aes(label = paste0(round(prop_screened * 100, 1), "%")),
    vjust = -0.3,
    size = 3.5
  ) +
  
  # Add dashed horizontal line for overall mean acceptance
  geom_hline(
    yintercept = overall_mean,
    linetype = "dashed",
    colour = "darkred",
    linewidth = 0.8
  ) +
  
  annotate(
    "text",
    x = -Inf,
    y = overall_mean,
    label = paste0("Mean: ",
                   round(overall_mean * 100, 1), "%"),
    hjust = 0,
    vjust = -0.5,
    colour = "darkred",
    size = 3.5
  ) +
  
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  
  labs(
    title = "Screening Acceptance Across Clinical Teams",
    x = "Clinical Team (Sample Size)",
    y = "Screening Acceptance (%)"
  ) +
  
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.x = element_text(margin = margin(t = 12)),
    axis.title.y = element_text(margin = margin(r = 12))
  )
```

## Statistical Modelling (05_statistical_modelling.R)

```{r statistical_modelling, echo=TRUE, eval=FALSE}
# ==========================================================
# Script: 05_statistical_modelling.R
# Purpose: Examine associations between patient
#          characteristics and breast cancer screening uptake
#          using univariate tests and multivariable logistic regression.
# Author: Nikolaos Samperis
# Date: 20/02/2026
# ==========================================================

# ----------------------------------------------------------
# 1. Load required libraries
# ----------------------------------------------------------
# tidyverse: data manipulation and visualisation
# broom: tidy model outputs for plotting
# car: variance inflation factor (VIF) for multicollinearity assessment

library(tidyverse)
library(broom)
library(car)

# -------------------------------------------------------------------------
# 2. Unadjusted Associations Between Patient Characteristics and Screening
# -------------------------------------------------------------------------

# Objective:
# Examine crude (unadjusted) associations between individual
# patient-level variables and screening uptake.
# Continuous variables are tested using t-tests (or Wilcoxon for skewed data),
# and categorical variables using chi-square tests.

# Null hypothesis (for each test):
# No association between the variable and screening uptake.

t.test(age ~ screening_decision, data = data)

# Non-parametric alternative due to right-skewed alcohol distribution
wilcox.test(alcohol_units ~ screening_decision, data = data)

chisq.test(table(data$employment_status, data$screening_decision))
chisq.test(table(data$n_dependents, data$screening_decision))

t.test(weight ~ screening_decision, data = data)
t.test(height ~ screening_decision, data = data)

# ------------------------------------------------------------------------
# 3. Multivariable Logistic Regression: Factors Associated with Screening
# ------------------------------------------------------------------------

# Objective: assess independent predictors of screening uptake while adjusting
# for key patient characteristics and clinical team. A parsimonious model is
# specified a priori based on plausibility and interpretability, then a more
# comprehensive model is explored as a sensitivity analysis.

# Set reference level for screening outcome (optional but recommended)
# Here, "N" (not screened) is treated as the reference category
data$screening_decision <- relevel(data$screening_decision, ref = "N")

# ---------------------------------
# 3.1 Primary (parsimonious) model
# ---------------------------------
# Core covariates were selected based on their plausible relationship with
# screening behaviour and potential to confound team comparisons (age,
# employment status, number of dependents, alcohol intake),
# alongside clinical team.

model_primary <- glm(
  screening_decision ~
    age +
    alcohol_units +
    employment_status +
    n_dependents +
    clinical_team,
  data = data,
  family = binomial
)

summary(model_primary)

# Convert log-odds coefficients to Odds Ratios with 95% CI
exp(cbind(
  OR = coef(model_primary),
  confint(model_primary)
))

# ---- Multicollinearity Check ----
# VIF > 5 indicates moderate concern
# VIF > 10 indicates serious multicollinearity

vif_primary <- vif(model_primary)
print(vif_primary)

# -------------------------------------------------------------------
# 3.2 Sensitivity analysis: extended model including weight & height
# -------------------------------------------------------------------
# Weight and height were associated with screening in univariate analyses.
# They are added here to assess whether these crude associations persist after
# adjustment, and whether inclusion materially changes the primary model estimates.

model_extended <- glm(
  screening_decision ~
    age +
    weight +
    height +
    alcohol_units +
    employment_status +
    n_dependents +
    clinical_team,
  data = data,
  family = binomial
)

summary(model_extended)

exp(cbind(
  OR = coef(model_extended),
  confint(model_extended)
))

# If the extended model shows inflated standard errors/unstable estimates
# (suggesting multicollinearity) and does not improve interpretability, the
# primary model is retained for final inference.

# ---- Multicollinearity Check (Extended Model) ----

vif_extended <- vif(model_extended)
print(vif_extended)

# ------------------------------------------------------------------------
# 3.3 Forest Plot – Adjusted Odds Ratios (Primary Model)
# ------------------------------------------------------------------------

# Tidy model output and exponentiate coefficients to OR scale
forest_data <- tidy(model_primary, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term_clean = case_when(
      term == "age" ~ "Age (per year increase)",
      term == "alcohol_units" ~ "Alcohol intake (per additional unit)",
      term == "n_dependents" ~ "Number of dependents",
      str_detect(term, "^employment_status") ~
        paste0(str_remove(term, "^employment_status"), " vs Employed"),
      str_detect(term, "^clinical_team") ~
        paste0(str_remove(term, "^clinical_team"),
               " vs DN01 (reference)"),
      TRUE ~ term
    ),
    group = if_else(str_detect(term, "^clinical_team"),
                    "Clinical Team",
                    "Patient Characteristics")
  )

# Order clinical teams numerically
teams_df <- forest_data %>%
  filter(group == "Clinical Team") %>%
  mutate(
    team_number = as.integer(str_extract(term_clean, "(?<=DN)\\d+"))
  ) %>%
  arrange(team_number) %>%
  mutate(term_clean = factor(term_clean, levels = rev(term_clean))) %>%
  select(-team_number)

# Order Patient Characteristics (keep logical order)
patients_df <- forest_data %>%
  filter(group == "Patient Characteristics") %>%
  mutate(term_clean = factor(term_clean, levels = rev(term_clean)))

# Combine
forest_data_final <- bind_rows(teams_df, patients_df)

# Forest plot of adjusted ORs (log scale)
ggplot(forest_data_final,
       aes(x = estimate, y = term_clean)) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.2,
                 linewidth = 0.8) +
  geom_point(size = 3) +
  geom_vline(xintercept = 1,
             linetype = "dashed",
             colour = "grey40") +
  scale_x_log10() +
  facet_wrap(~ group, scales = "free_y") +
  labs(
    x = "Adjusted Odds Ratio",
    y = "",
    title = "Factors Associated with Breast Cancer Screening Uptake",
    subtitle = "Values to the right of 1 indicate higher screening uptake"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, margin = margin(b = 10)),
    plot.subtitle = element_text(hjust = 0.5, margin = margin(b = 15)),
    strip.text = element_text(face = "bold"),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_text(margin = margin(t = 15))
  )
```






























